---
title: "Exploring synthetic HIV data"
description: |
  More figure examples.
author:
  - name: Brianna Argall
    url: {}
date: 2024-06-27
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
---

This dataset can be found at: 

Kuo, Nicholas (2022). The Heath Gym Synthetic HIV Dataset. figshare. Dataset. https://doi.org/10.6084/m9.figshare.19838470.v1

More information on the dataset is available at:
 Kuo et al. (2022). 
The Health Gym: Synthetic Health-Related Datasets for the Development of Reinforcement Learning Algorithms. arXiv preprint arXiv:2203.06369. 


```{r setup,include=FALSE}
knitr::opts_chunk$set(message = FALSE)

library(tidyverse)
library(knitr)

# Install package if necessary
if(!"ggridges" %in% installed.packages()) install.packages("ggridges")

library(ggridges)
```

```{r load_df, include=FALSE}
art_hiv <- read_csv("https://figshare.com/ndownloader/files/35249488", 
                    show_col_types = FALSE)
```


```{r variable_renaming}
art_hiv %>%
  rename_with(function(x) gsub(" ", "_", x)) %>%
  rename_with(function(x) gsub("\\(|\\)|\\.|-", "", x)) %>% # remove ().-
  mutate(Gender = factor(Gender, labels = c("1" = "Male", 
                                            "2" = "Female")),
         Ethnicity = factor(Ethnic, labels = c("2" = "African", 
                                               "3" = "Caucasian",
                                               "4" = "Other")),
         Base_Drug_Combo = factor(Base_Drug_Combo, 
                                  labels = c("0" = "FTC + TDF",
                                             "1" = "3TC + ABC",
                                             "2" = "FTC + TAF",
                                             "3" = "DRV + FTC + TDF",
                                             "4" = "FTC + RTVB + TDF",
                                             "5" = "Other")),
         Comp_INI = factor(Comp_INI, labels = c("0" = "DTG",
                                                "1" = "RAL",
                                                "2" = "EVG",
                                                "3" = "Not Applied")),
         Comp_NNRTI = factor(Comp_NNRTI, labels = c("0" = "NVP",
                                                    "1" = "EFV",
                                                    "2" = "RPV",
                                                    "3" = "Not Applied")),
         Extra_PI = factor(Extra_PI, labels = c("0" = "DRV",
                                                "1" = "RTVB",
                                                "2" = "LPV",
                                                "3" = "RTV",
                                                "4" = "ATV",
                                                "5" = "Not Applied")),
         Extra_pkEn = factor(Extra_pkEn, labels = c("0" = "False",
                                                    "1" = "True")),
         VL_M = factor(VL_M, labels = c("0" = "False",
                                        "1" = "True")),
         CD4_M = factor(CD4_M, labels = c("0" = "False",
                                          "1" = "True")),
         Drug_M = factor(Drug_M, labels = c("0" = "False",
                                            "1" = "True")),
         PatientID = factor(PatientID),
         Ethnic = NULL) %>%
  rename(Row_Number = "1") -> art_hiv

# Note: No observations list Asian ethnicity, hence this label was not used.
```



## Data dictionary

There are `r nrow(art_hiv)` rows and `r ncol(art_hiv)` columns in the dataframe.
A description of the columns is provided in this table. 
Although the description of the `Ethnicity` column includes Asian, 
no one was listed as having this ethnicity in the data.

```{r data_dictionary_table}
col_desc <- c("An integer indicating row number",
              "Viral Load (copies per mL)",
              "CD4 Absolute Count (copies per mm^3)",
              "CD4 Relative Count (copies per mm^3)",
              "Gender, male or female",
              "Base Drug Combination",
              "Complementary Integrase Inhibitor",
              "Complementary Non-Nucleoside Reverse Transcriptase Inhibitor",
              "Extra Protease Inhibitor",
              "Extra Pharmacokinetic Enhancer (y/n)",
              "Viral load measured (y/n)",
              "CD4 measured (y/n)",
              "Drug recorded (y/n)",
              "An integer indicating patient ID",
              "An integer indicating timepoints",
              "Ethnicity, Asian, African, Caucasian or other")

kable(cbind(lapply(art_hiv, class), col_desc), 
      col.names = c("Column", "Class", "Column description"))
```

## Univariate visualisations

```{r patient_row_amounts, include=FALSE}
art_hiv %>%
  group_by(PatientID) %>%
  tally() %>%
  distinct(n) -> rows_per_patient

art_hiv %>%
  distinct(PatientID) %>%
  tally() -> patient_amount

rows_per_patient * patient_amount == nrow(art_hiv) # TRUE :)
```


No visualisation is provided for `Row_Number`, `Timepoints` or for `Patient_ID`. 
There are `r rows_per_patient` rows per patient, each for a different point in time and `r patient_amount` patients.

### VL visualisation

```{r VL_basic_hist}
art_hiv %>%
  ggplot(aes(x = VL)) +
    geom_histogram() +
    scale_y_continuous(labels = scales::label_comma(),
                  name = "Count") +
    scale_x_continuous(labels = scales::label_comma(),
                 name = "Viral Load (copies per mL)") +
  ggtitle("Viral load histogram")
```


```{r VL_sqrt_both}
art_hiv %>%
  ggplot(aes(x = VL)) +
    geom_histogram() +
    scale_y_sqrt(labels = scales::label_comma(),
                  name = "Count",
                 # added breaks for low counts
                 breaks = c(10^3, 10^4, 5*10^4, 10^5, 1.5*10^5, 2*10^5)) +
    scale_x_sqrt(labels = scales::label_comma(),
                 breaks = c(100, 10^3, 5*10^3, 10^4, 2.5*10^4, 5*10^4, 7.5*10^4),
                 name = "Viral Load (copies per mL)") +
  ggtitle("Viral load histogram with square root transformation of both axes")
```

I chose to apply a square-root transformation to both axes to aid visibility of low values. The square-root transformation was chosen over a log transformation to change the apparent distribution shape less. 


```{r VL_timepoints_untrans}
art_hiv %>%
  filter(Timepoints %in% seq(0, 59, 8)) %>%
  ggplot(aes(x = VL)) +
    geom_histogram(binwidth = 5000) +
    scale_y_continuous(labels = scales::label_comma(),
                       name = "Count") +
    scale_x_continuous(labels = scales::label_comma(), 
                       name = "Viral Load (copies per mL)") +
  facet_wrap(.~factor(Timepoints), ncol = 3) +
  ggtitle("Viral load histograms for selected timepoints")
```

```{r VL_timepoints_sqrt}
art_hiv %>%
  filter(Timepoints %in% seq(0, 59, 8)) %>%
  ggplot(aes(x = VL)) +
    geom_histogram() +
    scale_y_sqrt(labels = scales::label_comma(),
                 # default breaks overlapped
                 breaks = c(10, 250, 10^3, 2.5*10^3, 5*10^3),
                 name = "Count") +
    scale_x_sqrt(labels = scales::label_comma(), 
                 name = "Viral Load (copies per mL)",
                 # default breaks overlapped
                 breaks = c(100, 5*10^3, 2.5*10^4, 7.5*10^4)) +
  facet_wrap(.~factor(Timepoints), ncol = 3) +
  ggtitle("Viral load histograms for selected timepoints with transformed axes")
```

Histograms of viral load displayed for selected time points. A square root transformation has been applied to each axis. This aids visibility of low values and facilitates comparison with the main histograms for viral load.

There is a ridge plot of viral load presented near the end of the document, placed there due to its length.


### CD4 visualisation

```{r}
art_hiv %>%
  ggplot(aes(x = CD4)) +
  geom_histogram() +
  ylab("Count") +
  xlab("CD4+") +
  ggtitle("Absolute CD4+ histogram") +
  scale_y_continuous(labels = scales::label_comma())

```

```{r}
art_hiv %>%
  ggplot(aes(x = CD4)) +
  geom_histogram() +
  ylab("Count") +
  xlab("CD4+") +
  scale_x_sqrt(breaks = c(100, 400, 1000, 2500, 5000, 10000, 15000)) +
  scale_y_sqrt(labels = scales::label_comma(),
               breaks = c(10^2, 2.5*10^3, 10^4, 3*10^4, 5*10^4, 10^5)) +
  ggtitle("Viral load histogram with square root transformation of both axes")
```


### Rel CD4 visualisation

```{r}
art_hiv %>%
  mutate(high_rel = if_else(Rel_CD4 >= 100, "high", "expected")) %>%
  ggplot(aes(x = Rel_CD4, fill = high_rel)) +
  geom_histogram() +
  geom_vline(xintercept = 100, linetype = "longdash", alpha = 0.25) +
  ggtitle("Relative CD4+ histogram") +
  xlab("Relative CD4+") +
  scale_fill_manual("Rel_CD4 >= 100?", 
                    values = c("#1b9e77", "#d95f02"),
                    labels = c("No", "Yes"))
```
```{r}
art_hiv %>%
  mutate(high_rel = if_else(Rel_CD4 >= 100, "high", "expected")) %>%
  ggplot(aes(x = Rel_CD4, fill = high_rel)) +
  # I tried changing bin or bin_width to have all >= 100 on one side but failed
  geom_histogram(bins = 20) +
  geom_vline(xintercept = 100, linetype = "longdash", alpha = 0.25) +
  ggtitle("Relative CD4+ histogram with square root transformation to both axes") +
  scale_x_sqrt(breaks = c(5, 25, 50, 75, 100, 200, 300, 400),
               name = "Relative CD4+") +
  scale_y_sqrt(labels = scales::label_comma(),
               name = "Count",
               breaks = c(100, 10^3, 10^4, 3*10^4, 6*10^4, 9*10^4)) +
  # this would be neater if I used LaTex for the symbol
  scale_fill_manual("Rel_CD4 >= 100?", 
                    values = c("#1b9e77", "#d95f02"),
                    labels = c("No", "Yes"))
```


### Gender visualisation

```{r}
art_hiv %>%
  group_by(PatientID, Gender) %>%
  filter(row_number(PatientID) == 1) %>%
  ggplot(aes(x = Gender, fill = Gender)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma(),
                     name = "Count",
                     # This fixes an issue with text being clipped at the top
                     expand = expansion(mult = 0.1)) +
  scale_fill_manual(values = c("#8dd3c7", "#bebada")) +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None") +
  ggtitle("Patient gender bar chart") +
  # vjust here shifts the text fully on top of the bar
  stat_count(geom = "text", aes(label = after_stat(count)), vjust = -0.5)
```

### Base Drug Combo visualisation

```{r}
art_hiv %>%
  # Not sure how to programmatically reorder according to size
  # while keeping "Other" on the end, so manual it is
  mutate(Base_Drug_Combo = factor(Base_Drug_Combo, 
                                  levels = c("FTC + TDF", 
                                             "3TC + ABC", 
                                             "DRV + FTC + TDF",
                                             "FTC + RTVB + TDF",
                                             "FTC + TAF",
                                             "Other"))) %>%
  ggplot(aes(x = Base_Drug_Combo, fill = Base_Drug_Combo)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma(),
                     name = "Count") +
  # Without modification the labels overlap
  scale_x_discrete(labels = gsub("\\s", "\n", levels(art_hiv$Base_Drug_Combo)),
                   name = "Base Drug Combination") +
  # I tried to put a little bit of thought into the colours
  # e.g. 3TC & ABC don't appear elsewhere so no blue tie in
  # I kept these colours for all categorical >2 variables
  scale_fill_manual(values = c("#b3e2cd", "#fdcdac", "#cbd5e8", 
                               "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None") +
  ggtitle("Base drug combination bar chart")

```

Note that this is per observation/time point, not per person. 

### Comp INI visualisation

```{r}
art_hiv %>%
  ggplot(aes(x = reorder(Comp_INI, Comp_INI, length), fill = Comp_INI)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = c("#b3e2cd", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  ggtitle( "Complementary integrase inhibitor bar chart") +
  xlab("Complementary integrase inhibitor") +
  ylab("Count") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```

Note that this is per observation/time point, not per person. 

```{r}
art_hiv %>%
  filter(Comp_INI != "Not Applied") %>%
  ggplot(aes(x = reorder(Comp_INI, Comp_INI, length), fill = Comp_INI)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma(),
                     # Added 2,500 break to defaults
                     breaks = c(2.5*10^3, 2*10^4, 4*10^4, 6*10^4)) +
  scale_fill_manual(values = c("#b3e2cd", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  ggtitle( "Complementary integrase inhibitor bar chart, not showing unused") +
  xlab("Complementary integrase inhibitor") +
  ylab("Count") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```

### Comp NNRTI visualisation

```{r}
art_hiv %>%
  ggplot(aes(x = Comp_NNRTI, fill = Comp_NNRTI)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = c("#b3e2cd", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  ggtitle( "Complementary NNRTI bar chart") +
  xlab("Complementary NNRTI") +
  ylab("Count") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```

Note that this is per observation/time point, not per person. 

```{r}
art_hiv %>%
  ggplot(aes(x = Comp_NNRTI, fill = Comp_NNRTI)) +
  geom_bar() +
  scale_y_log10(labels = scales::label_comma()) +
  scale_fill_manual(values = c("#b3e2cd", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  ggtitle( "Complementary NNRTI bar chart using log scale") +
  xlab("Complementary NNRTI") +
  ylab("Count") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```


### Extra PI visualisation

```{r}
art_hiv %>%
  ggplot(aes(x = reorder(Extra_PI, Extra_PI, length), fill = Extra_PI)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma(),
                     name = "Count") +
  xlab("Extra Protease Inhibitor") +
  ggtitle("Extra protease inhibitor bar chart") +
  scale_fill_manual(values = c("#b3e2cd", "#fdcdac", "#cbd5e8", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```

Note that this is per observation/time point, not per person.

```{r}
art_hiv %>%
  filter(Extra_PI != "Not Applied") %>%
  ggplot(aes(x = reorder(Extra_PI, Extra_PI, length), fill = Extra_PI)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma(),
                     name = "Count") +
  xlab("Extra Protease Inhibitor") +
  scale_fill_manual(values = c("#b3e2cd", "#fdcdac", "#cbd5e8", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None") +
  ggtitle("Extra protease inhibitor bar chart, not showing unused")
```


### Extra pk En visualisation

```{r}
binary_figure <- function(xvalue) {
  art_hiv %>%
    ggplot(aes(x = {{xvalue}}, fill = {{xvalue}})) +
    geom_bar() +
    scale_y_continuous(labels = scales::label_comma(),
                       name = "Count",
                       expand = expansion(mult = 0.1)) +
    # I just think the only grey is boring & defaults were ugly (too saturated)
    # I also wanted to avoid red or green symbolism
    scale_fill_manual(values = c("#8dd3c7", "#bebada")) +
    theme_bw() +
    theme(panel.border = element_blank(),
          panel.grid.major.x = element_blank(),
          legend.position = "None") +
    stat_count(geom = "text", aes(label = after_stat(count)), vjust = -0.5)
}

```

```{r}
binary_figure(Extra_pkEn) +
  xlab("Extra Pharmacokinetic Enhancer provided") +
  ggtitle("Extra pharmacokinetic enhancer bar chart")
```

Note that this is per observation/time point, not per person. 


### VL M visualisation
```{r}
binary_figure(VL_M) +
  xlab("Viral load measurement") +
  ggtitle("Viral load measurement bar chart")
```

Note that this is per observation/time point, not per person. 

### CD4 M visualisation

```{r}
binary_figure(CD4_M) +
  xlab("CD4+ measurement") +
  ggtitle("CD4+ measurement bar chart")
```

Note that this is per observation/time point, not per person. 

### Drug M visualisation

```{r}
binary_figure(Drug_M) +
  xlab("Drug measurement/recorded") +
  ggtitle("Drug measurement bar chart")
```


Note that this is per observation/time point, not per person. 

### Ethnicity visualisation


```{r}
art_hiv %>%
  group_by(PatientID, Ethnicity) %>%
  filter(row_number(PatientID) == 1) %>%
  ggplot(aes(x = Ethnicity, fill = Ethnicity)) +
  geom_bar() +
  scale_y_continuous(labels = scales::label_comma(),
                     expand = expansion(mult = 0.1)) +
  scale_fill_manual(values = c("#b3e2cd", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  ggtitle( "Ethnicity bar chart") +
  ylab("Count") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None") +
  stat_count(geom = "text", aes(label = after_stat(count)), vjust = -0.5)
```

```{r}
art_hiv %>%
  group_by(PatientID, Ethnicity) %>%
  filter(row_number(PatientID) == 1) %>%
  ggplot(aes(x = Ethnicity, fill = Ethnicity)) +
  geom_bar() +
  scale_y_log10(labels = scales::label_comma(),
                expand = expansion(mult = 0.1)) +
  scale_fill_manual(values = c("#b3e2cd", "#f4cae4", "#e6f5c9", "#E1E1E1")) +
  ggtitle( "Ethnicity bar chart with log scale") +
  ylab("Count") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```

## Ridgeline plot


```{r, fig.height=12}
art_hiv %>%
  mutate(Timepoints = as.factor(Timepoints)) %>%
  ggplot(aes(x = VL, y = Timepoints, fill = after_stat(x))) +
  geom_density_ridges_gradient() +
  scale_x_log10(labels = scales::label_comma(),
                breaks = c(10, 10^2, 10^3, 10^4, 10^5),
                name = "Viral Load",
                # 2nd axis aids interpretation given plot height
                sec.axis = dup_axis()) +
  # I want readers to move forwards through time as they scroll down
  scale_y_discrete(limits = rev) +
  scale_fill_viridis_c(trans = "log") +
  theme_bw() +
  theme(panel.border = element_blank(),
        legend.position = "None")
```

## Other exploration / data checking

```{r class.source = "fold-show"}
# confirming that Row_Number is just row numbers and not some other sequence
art_hiv %>%
  mutate(row_n = row_number(Row_Number),
         consistent = ifelse(row_n - 1  == Row_Number, TRUE, FALSE)) %>%
  select(consistent) %>%
  summary()
  
```
  
As you can see below, even where a binary indicator shows a variable as not being measured this does not mean there is no value provided for that variable.
  
This may be an artifact from the data being synthetic. 

```{r class.source = "fold-show"}
art_hiv %>%
  filter(VL_M == "False") %>%
  select(VL, VL_M) %>%
  head()
```


```{r class.source = "fold-show"}
art_hiv %>%
  filter(CD4_M == "False") %>%
  select(CD4, Rel_CD4, CD4_M) %>%
  head()
```

```{r class.source = "fold-show"}
art_hiv %>%
  filter(Drug_M == "False") %>%
  select(Drug_M, Base_Drug_Combo, Comp_INI, Comp_NNRTI, Extra_PI, Extra_pkEn) %>%
  head()
```



